Towards a Verified Tableau Prover for an Unquantified Fragment of Set Theory

### 2023-02-23 paper/CADE/paper.tex
Continue

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is an unquantified fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  \keywords{First TODO \and Second TODO\todo{keywords}}
\end{abstract}

@@

\subsection{Contributions}
This paper is about the formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

Abstract framework to prove completeness for first-order tableau systems~\cite{completeness_coinductive}.
Verified sequent calculus prover for first-order logic~\cite{sequent_first_order}.
Tableau calculus for temporal logic CTL in Coq~\cite{ctl_tableau}
Abstract framework for synthetic completeness in AFP~\cite{synthetic_completeness_afp}.

While a lot of work has gone into the verification of decision procedures, seemingly little has been put towards the verification of tableau calculi.
In Lean, there is a verified tableau calculus for basic modal logic~\cite{modal_logic_lean}.
In Coq, there is a formalisation of a tableau calculus for ecumenical logic~\cite{ecumenical_logic}; it is, however, missing a completeness proof.
Concerning Isabelle/HOL, there is formalisation of a sound and complete tableau calculus for hybrid logic~\cite{hybrid_logic} available in the \textit{Archive of Formal Proofs}~\cite{hybrid_logic_afp}.
The termination proof is part of ongoing work~\cite{hybrid_logic_workshop}.
% TODO: Formalisation de la logique temporelle dans le coq.

@@

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
We will often drop the atom constructor \lstinline!&\fmAtom{}&! to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.

Like the function \lstinline!vars!, the automatically defined function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! retrieves all set atoms in a formula.
We combine these functions to extract all the variables occurring in a set formula.
\begin{lstlisting}
definition vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!, which returns all the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the usual way (cf. Appendix \todo{reference Appendix}).

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \textit{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \textit{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ --- that is, the axioms of ZF --- except for the axiom of infinity.\todo{Membership is well-founded.}
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions 
\begin{itemize}
  \item \lstinline!&\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf! and 
  \item \lstinline!&\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf!
\end{itemize}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to the appendix.

We write \lstinline!M $\models$ $\phi$! for the judgement that the formula $\phi$ holds under the valuation function \lstinline!M!.
The implementation of $\models$ coincides with the interpretation function of \citeauthor{lqe}~\cite{lqe}.
As usual, a formula \lstinline!$\phi$! is called \textit{satisfiable} if there exists a model \lstinline!M! with \lstinline!M $\models$ $\phi$!.
Otherwise, \lstinline!$\phi$! is called \textit{unsatisfiable}.
\todo[inline]{Introduce the initial formula}

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulae are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau. 
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

In the standard tableau calculus for propositional logic as \citeauthor{tableau}~\cite{tableau} describes it, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For \MLSS{}, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\lstinline!t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$!.

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \textit{closed} if all of its branches are closed.

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!$\interS$! && Rules for \lstinline!$\diffS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{c}
      $\vdots$ \\
    \end{tabular}
    & &
    \begin{tabular}{c}
      $\vdots$ \\
    \end{tabular}
    \\
    \\

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set terms of \lstinline!l!, i.e.\ it occurs directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

The more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\lstinline!s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$!,
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

Due to boundedness, repeated expansion with linear rules eventually results in a \textit{linearly saturated} branch, i.e.\ a branch where no application of linear rules would produce new formulae.
\begin{lstlisting}
definition lin_sat b $\equiv$ $\forall$b'. &\lexpands{b'}{b}& $\longrightarrow$ set b' $\subseteq$ set b
\end{lstlisting}

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here.}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \textit{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \textit{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

A linearly saturated branch where no further branching is possible is called a \textit{saturated} branch. 
\begin{lstlisting}
definition sat b $\equiv$ lin_sat b $\land$ ($\nexists$bs'. &\fexpands{bs'}{b}&)
\end{lstlisting}

Note that even branching rules are defined such that they never create new subterms with exception of the last rule, which adds a new variable to the branch.
These variables serve to manifest an inequality; hence, we call them \textit{witnesses}.
\begin{lstlisting}
definition wits b $\equiv$ vars b - vars (last b)
\end{lstlisting}

>>>>

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is an unquantified fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  Furthermore, we extend the calculus with a light-weight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{First TODO \and Second TODO\todo{keywords}}
\end{abstract}

@@

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

There is a large body of work at the intersection of ITPs and tableau methods but to keep with the theme of the paper we only consider formalisations of correctness here.
For first-order logic, there are abstract completeness proofs using the \emph{Beth-Hintikka style} of possibly infinite derivation trees~\cite{completeness_coinductive} as well as the \emph{Henkin style} of maximally consistent sets~\cite{synthetic_completeness_afp}.
Both are abstract enough to be instantiated with a wide range of concrete calculi.
A more concrete formalisation~\cite{sequent_first_order} verifies a sequent calculus for first-order logic whose completeness proof is via a translation to semantic tableau.

Beyond completeness, we target decidability which is more attainable for propositional logic.  
There is a verified tableau calculus for the modal logic S5~\cite{lean_modal_logic} in Lean and one for hybrid logic~\cite{hybrid_logic} in Isabelle/HOL.
Both of these formalisation do not prove termination but there is a formalisation of a tableau calculus for the temporal logic CTL in Coq~\cite{ctl_tableau} that does.
% TODO: Formalisation de la logique temporelle dans le coq.

@@

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
We will often drop the atom constructor \lstinline!&\fmAtom{}&! to reduce clutter.
Additionally, we use \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.

Similarly to \lstinline!vars!, we get the function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! for free which retrieves all set atoms in a formula.
We combine these functions to extract all the variables occurring in a set formula.
\begin{lstlisting}
definition vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!.
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the usual way (cf.\ \appendixname~\ref{appendix:subterms}).

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \emph{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \emph{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ --- that is, the axioms of ZF --- except for the axiom of infinity.
In particular, the membership relation is well-founded.
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions 
\begin{itemize}
  \item \lstinline!&\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf! and 
  \item \lstinline!&\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf!
\end{itemize}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to formalisation (cf.\ \appendixname~\ref{appendix:semantics}).

We write \lstinline!M $\models$ $\phi$! for the judgement that the formula $\phi$ holds under the valuation function \lstinline!M!.
The implementation of $\models$ coincides with the interpretation function of \citeauthor{lqe}~\cite{lqe}.
As usual, a formula \lstinline!$\phi$! is called \emph{satisfiable} if there exists a model \lstinline!M! with \lstinline!M $\models$ $\phi$!.
Otherwise, \lstinline!$\phi$! is called \emph{unsatisfiable}.

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulae are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau.
We sometimes call this formula the \emph{initial formula}.
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

In the standard tableau calculus for propositional logic as \citeauthor{tableau}~\cite{tableau} describes it, a branch is called \emph{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \emph{open} if it is not closed.
For \MLSS{}, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\lstinline!t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$!.

@@

abbreviation bopen b $\equiv$ $\neg$ bclosed b
\end{lstlisting}
A tableau is called \emph{closed} if all of its branches are closed.

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

    Rules for \lstinline!$\interS$! && Rules for \lstinline!$\diffS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{c}
      $\vdots$
    \end{tabular}
    & &
    \begin{tabular}{c}
      $\vdots$ \\
    \end{tabular}
    \\[0.9em]

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands} We omit the rules for $\interS$ and $\diffS$ as they coincide with the original paper (cf.\ \appendixname~\ref{appendix:rules}).}
\end{table*}

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set terms of \lstinline!l!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

The more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\lstinline!s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$!,
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:correct}.

Due to boundedness, repeated expansion with linear rules eventually results in a \emph{linearly saturated} branch, i.e.\ a branch where no application of linear rules would produce new formulae.
\begin{lstlisting}
definition lin_sat b $\equiv$ $\forall$b'. &\lexpands{b'}{b}& $\longrightarrow$ set b' $\subseteq$ set b
\end{lstlisting}

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. The omitted rules coincide with the original paper (cf.\ \appendixname~\ref{appendix:rules}).}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \emph{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \emph{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

A linearly saturated branch where no further branching is possible is called a \emph{saturated} branch. 
\begin{lstlisting}
definition sat b $\equiv$ lin_sat b $\land$ ($\nexists$bs'. &\fexpands{bs'}{b}&)
\end{lstlisting}

Note that even branching rules are defined such that they never create new subterms with exception of the last rule, which adds a new variable to the branch.
These variables serve to manifest an inequality; hence, we call them \emph{witnesses}.
\begin{lstlisting}
definition wits b $\equiv$ vars b - vars (last b)
\end{lstlisting}

\section{A Decision Procedure for MLSS}
The mechanics of the decision procedure are typical for a procedure based on a tableau calculus:
it decides the satisfiability of a given formula \lstinline!$\phi$! by determining whether the formula has a closed tableau.
More specifically, it initialises the tableau with the singleton branch \lstinline![$\phi$]! and checks whether this branch can be expanded to a closed tableau.
